---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Ben Waggener"
date: "2/07/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

"What are the dimensions of the Denny’s dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?"

```{r ex1_dannysdataset_dimensions}

glimpse(dennys)

nrow(dennys)
ncol(dennys)

```
There are 1,643 rows in the Dennys dataset which represents individual Dennys locations. There are 6 columns which represents the variables that describe the location.

### Exercise 2

"What are the dimensions of the La Quinta’s dataset? What does each row in the dataset represent? What are the variables?"

``` {r ex2_laquinta_dim}

glimpse(laquinta)

nrow(laquinta)
ncol(laquinta)


```
There are 909 rows in the La Quinta dataset which represent individual La Quintas. There are 6 columns that represent the variables that describe the location of each hotel.


### Exercise 3

Looking at the webpages, I saw that there are no Dennys outside of the United States while there are La Quintas outside of the United States.

This includes locations Canada, Mexico, China, New Zealand, Turkey, UAE, Chile, Colombia, and Ecuador.
...

### Exercise 4

Filtering using the state variable might be the best way to tell if there are locations outside of the US because as far as I can tell these are unique to America. For a rough method maybe you could set a boundary with the lines of longitude and latitude but I don't know if that would be precise enough to capture the entire United states and not accidentally include hotels in Mexico or Canada.

...

### Exercise 5

"Find the Denny’s locations that are outside the US, if any. To do so, filter the Denny’s locations for observations where state is not in states$abbreviation. The code for this is given below. Note that the %in% operator matches the states listed in the state variable to those listed in states$abbreviation. The ! operator means not. Are there any Denny’s locations outside the US?"

``` {r ex5_dennysloc}

dennys %>%
  filter(!(state %in% states$abbreviation))

```

...

### Exercise 6

...

 
Add exercise headings as needed.

